---
title: "The State of Auckland's Buses"
author: "A visual overview of the performance of buses in the Auckland transport network"
date: "Based on historical and real-time data, updated approximately every 30 minutes,<br>obtained from Auckland Transport's public GTFS API"
output: 
    revealjs::revealjs_presentation:
        transition: fade
        css: march.css
        reveal_options:
            controls: false
            loop: true
            autoSlide: 30000
---


```{r setup, echo=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, error=FALSE)
```

## Real-time and historical bus data

```{r load_stuff, results='hide'}
library(lubridate)
source('makeplots.R')
load('~/Dropbox/gtfs/delayhistory.rda')
palfun <- ggthemes::colorblind_pal()
colpal <- list(late = palfun(4)[2],
               ontime = palfun(4)[4],
               early = palfun(4)[3])
palfun2 <- function() viridis::viridis(10)[c(9, 7, 5)]
colpal2 <- list(q50 = palfun2()[1],
                q75 = palfun2()[2],
                q90 = palfun2()[3])
```

<style>
.reveal section .late {
    color: `r colpal$late`;
}
.reveal section .ontime {
    color: `r colpal$ontime`;
}
.reveal section .early {
    color: `r colpal$early`;
}
.reveal section .q90 {
    color: `r colpal2$q90`;
}
.reveal section .q75 {
    color: `r colpal2$q75`;
}
.reveal section .q50 {
    color: `r colpal2$q50`;
}
</style>


- Using the public GTFS API provided by Auckland Transport, we obtain &hellip;
    - <strong>GPS positions</strong> (= position on a map), and
    - the <strong>current delay</strong> (= difference between scheduled and actual arrival time at last stop)  

    &hellip; for all active buses in the network.

<div></div><br>

- Using data collected over the last 10 weeks, we can see if today is better or worse than usual

## How many buses?

```{r ngraph,fig.height=3,fig.width=7}
Nhist %>% filter(date == "2018-02-19") %>%
    ggplot(aes(x = time)) + 
        geom_area(aes(y = vlate), fill = colpal$late) +
        geom_area(aes(y = ontime), fill = colpal$ontime) + 
        geom_area(aes(y = early), fill = colpal$early) +
        xlim(5, 24) +
        xlab("Time (hour)") + ylab("Number of buses") +
        geom_label(aes(x = 8.25, y = 950, label = 'A')) +
        geom_label(aes(x = 15.75, y = 850, label = 'B')) +
        geom_label(aes(x = 17.75, y = 850, label = 'C'))
```

* Peak periods: __A__ morning, __B__ school, __C__ evening
* The coloured areas represent the proportion<br> of buses that are, from top to bottom,
    - <span class='late'>late by more than 5 minutes</span>
    - <span class='ontime'>on time (within 5 minutes of schedule)</span>
    - <span class='early'>more than 5 minutes early</span>


```{r latest_ontime}
Nlatest <- Nhist[nrow(Nhist), ]
Qlatest <- Qhist[nrow(Qhist), ]
tcur <- Nlatest$time
tlatest <- paste0(floor(tcur) - ifelse(tcur >= 13, 12, 0), 
                  ":", sprintf("%02d", round((tcur %% 1) * 60)), " ",
                  ifelse(floor(tcur) < 12, "am", "pm"))
total <- Nlatest$vlate
early <- Nlatest$early
ontime <- Nlatest$ontime - early
late <- total - ontime - early
```
## As of `r tlatest` today


* <span class='bigN ontime'>`r ontime`</span> of <span class='bigN total'>`r total`</span> buses were on time
* <span class='bigN late'>`r late`</span> were running more than 5 minutes late
* <span class='bigN early'>`r early`</span> were more than 5 minutes early

```{r todaytrace,fig.height=3,fig.width=10}
Nday <- Nhist %>% 
    filter(Nhist$date == Sys.Date())
gridExtra::grid.arrange(
    ggplot(Nday, aes(x = time)) + 
        geom_area(aes(y = vlate), fill = colpal$late) +
        geom_area(aes(y = ontime), fill = colpal$ontime) + 
        geom_area(aes(y = early), fill = colpal$early) +
        xlim(5, 24) +
        xlab("Time (hour)") + ylab("Number of buses"),
    ggplot(Nday, aes(x = time)) + 
        geom_path(aes(y = (vlate - ontime) / vlate * 100),
            color = colpal$late, lwd = 2) +
        geom_path(aes(y = (ontime - early) / vlate * 100),
            color = colpal$ontime, lwd = 2) + 
        geom_path(aes(y = early / vlate * 100),
            color = colpal$early, lwd = 2) +
        ylim(0, 100) + xlim(5, 24) +
        xlab("Time (hour)") + ylab("Percent of buses"),
    ncol = 2)
```

## Today vs. last 10 `r format(today(), '%A')`s

* number and percentage of buses <strong class='early'>early</strong>, 
  <strong class='ontime'>ontime</strong>, or <strong class='late'>late</strong>
```{r numbersHist, fig.height=2.8, fig.width=10}
Nyear <- Nhist %>%
    filter(date >= today() - weeks(10) & date < today()) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday', 'Saturday', 'Sunday')))
## Filter out Mon-Fri with max less than 400
dateMax <- with(Nyear, tapply(vlate, date, max))
#Nyear <- Nyear %>%
#    filter(dateMax[Nyear$date %>% as.character] > 730)
Nlatest$dow <- factor(format(Nlatest$date, "%A"), levels = levels(Nyear$dow))
Nday$dow <- factor(format(Nday$date, "%A"), levels = levels(Nyear$dow))

gridExtra::grid.arrange(
    ggplot(Nyear %>% filter(dow == format(Sys.Date(), '%A')), 
           aes(x = time, group = date)) + 
        geom_line(aes(y = vlate), color = colpal$late) +
        geom_line(aes(y = ontime), color = colpal$ontime) +
        geom_line(aes(y = early), color = colpal$early) +
        xlab("Time (hour)") + ylab("Number of buses"),
    ggplot(Nyear %>% filter(dow == format(Sys.Date(), '%A')), 
           aes(x = time, group = date)) + 
        geom_line(aes(y = (vlate - ontime) / vlate * 100), color = colpal$late) +
        geom_line(aes(y = (ontime - early) / vlate * 100), color = colpal$ontime) +
        geom_line(aes(y = early / vlate * 100), color = colpal$early) +
        xlab("Time (hour)") + ylab("Percent of buses") + ylim(0, 100),
    nrow = 1
)
```
- comparing __average__ (dashed) of past Thursdays with today
```{r averages, fig.height=2.8, fig.width=10}
gridExtra::grid.arrange(
    ggplot(Nyear %>% filter(dow == format(Sys.Date(), '%A')), aes(x = time)) + 
        geom_area(aes(y = vlate), data = Nday, 
            fill=colpal$late, size = 2, stroke=2, pch = 4, alpha = 0.6) +
        geom_area(aes(y = ontime), data = Nday, 
            fill=colpal$ontime, size = 2, stroke=2, pch= 4, alpha = 0.6) +
        geom_area(aes(y = early), data = Nday, 
            fill=colpal$early, size = 2, stroke=2, pch = 4, alpha = 0.6) +
        geom_smooth(aes(y = vlate), 
            span = 0.1, method='loess', se=FALSE, color=colpal$late, lty = 2) +
        geom_smooth(aes(y = ontime), 
            span = 0.1, method='loess', se=FALSE, color=colpal$ontime, lty = 2) +
        geom_smooth(aes(y = early), 
            span = 0.1, method='loess', se=FALSE, color=colpal$early, lty = 2) +
        xlab("Time (hour)") + ylab("Number of buses"),
    ggplot(Nyear %>% filter(dow == format(Sys.Date(), '%A')), aes(x = time)) + 
        geom_path(aes(y = (vlate - ontime) / vlate * 100), data = Nday,
            color=colpal$late, size = 2, stroke=2, pch = 4, alpha = 0.7) +
        geom_path(aes(y = (ontime - early) / vlate * 100), data = Nday, 
            color=colpal$ontime, size = 2, stroke=2, pch = 4, alpha = 0.7) +
        geom_path(aes(y = early / vlate * 100), data = Nday, 
            color=colpal$early, size = 2, stroke=2, pch = 4, alpha = 0.7) +
        geom_smooth(aes(y = (vlate - ontime) / vlate * 100), 
            span = 0.1, method='loess', se=FALSE, color=colpal$late, lty = 2) +
        geom_smooth(aes(y = (ontime - early) / vlate * 100), 
            span = 0.1, method='loess', se=FALSE, color=colpal$ontime, lty = 2) +
        geom_smooth(aes(y = early / vlate * 100), 
            span = 0.1, method='loess', se=FALSE, color=colpal$early, lty = 2) +
        xlab("Time (hour)") + ylab("Percent of buses") + ylim(0, 100),
    nrow = 1
)
```


## This week vs. last 4 weeks

```{r mmsofar, fig.height=5.5, fig.width=9}
Monday <- today() - days(as.numeric(format(today(), "%u")) - 1)
mm <- Nhist %>%
    filter(date >= Monday - weeks(4) & date <= today()) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday', 'Saturday', 'Sunday')),
           week = (format(date, '%W') %>% as.numeric) - format(Monday, '%W') %>% as.numeric) %>%
    select(zero, vearly, early, ontime, late, vlate, time, dow, week) %>%
    mutate(week = factor(week, levels = -4:-0,
                         labels = c(paste(4:2, 'weeks ago'), "Last week", "This week")))

ggplot(mm, aes(x = time)) +
    geom_path(aes(y = (vlate - ontime) / vlate * 100), lwd = 1,
        color = colpal$late, data = mm) + 
    geom_path(aes(y = (ontime - early) / vlate * 100), lwd = 1,
        color = colpal$ontime, data = mm) + 
    geom_path(aes(y = early / vlate * 100), lwd = 1,
        color = colpal$early, data = mm) + 
    facet_grid(week~dow, drop = FALSE) +
    xlab("Time (hour)") + ylab("Percent of buses") + ylim(0, 100)
```

* percentage of buses <strong class='early'>early</strong>, 
<strong class='ontime'>ontime</strong>, or <strong class='late'>late</strong>


## So, just how late is late?

```{r}
library(RProtoBuf)
readProtoFiles(dir = "../assets/protobuf")
pb <- read(transit_realtime.FeedMessage, '~/Dropbox/gtfs/trip_updates.pb')
tobs <- as.POSIXct(pb$header$timestamp, origin = "1970-01-01")
```

* Each dot represents one vehicle's delay at `r format(tobs, '%l:%M %p')`
```{r nowdelays, fig.height=5, fig.width=10}
delays <- data.frame(
    delay = sapply(pb$entity, function(x) {
        stu <- x$trip_update$stop_time_update[[1]]
        if (stu$has('arrival')) return (stu$arrival$delay)
        return (stu$departure$delay)
    })) %>%
    filter(delay > -60*60 & delay < 60*60) %>%
    mutate(g = rep("delay", n()),
           delay = pmax(-20, pmin(30, delay / 60)))
p1 <- ggplot(delays, aes(x = delay)) + 
    geom_dotplot(binwidth = 1/3) + 
    xlim(-20, 30) +
    theme(axis.ticks.y = element_blank(),
          axis.text.y = element_blank()) +
    xlab("Delay (minutes)") + ylab("")
delayq <- data.frame(
        q1 = quantile(delays$delay, 0.05),
        q2 = quantile(delays$delay, 0.125),
        q3 = quantile(delays$delay, 0.25),
        q4 = quantile(delays$delay, 0.75),
        q5 = quantile(delays$delay, 0.875),
        q6 = quantile(delays$delay, 0.95)
    )
p2 <- ggplot(delayq, aes(xmin = -1, xmax = 1)) +
    geom_rect(aes(ymin = q1, ymax = q6), fill = colpal2$q90) +
    geom_rect(aes(ymin = q2, ymax = q5), fill = colpal2$q75) +
    geom_rect(aes(ymin = q3, ymax = q4), fill = colpal2$q50) +
    coord_flip() +
    ylim(-20, 30) + xlim(-2, 2) +
    theme_void() +
    ylab("") + xlab("")
gridExtra::grid.arrange(p1, p2, heights = c(6, 1))
```
* the bar contains 
<span class='q90'>90%</span>,
<span class='q75'>75%</span>, and
<span class='q50'>50%</span> of buses


## Delays so far today

```{r delayToday,fig.height=5,fig.width=10,fig.align='center'}
Qday <- Qhist %>% 
    filter(date == Sys.Date())
 ggplot(Qday, aes(x = time)) +
    geom_linerange(aes(ymin = q5, ymax = q95), color = colpal2$q90, lwd = 4) + 
    geom_linerange(aes(ymin = q125, ymax = q875), color = colpal2$q75, lwd = 4) + 
    geom_linerange(aes(ymin = q25, ymax = q75), color = colpal2$q50, lwd = 4) + 
    xlab("Time (hour)") + ylab("Delay (minutes)") + 
    xlim(min(Qhist$time), max(Qhist$time)) + ylim(-15, 30)
```

* rotating the bar from the previous graph, we see how the distribution of delays has changed today


## Today vs. last 10 `r format(today(), '%A')`s

```{r todayQ, fig.height=5, fig.width=10, fig.align='center'}
Qweeks <- Qhist %>%
    filter(date >= today() - weeks(10) & date < today()) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday', 'Saturday', 'Sunday')))
Qdow <- Qweeks %>% filter(dow == format(today(), "%A"))
halpha <- 0.5
ggplot(Qdow, aes(x = time)) +
    geom_line(aes(y = q5, group = date),
        color=colpal2$q90, lty = 4, lwd = 0.5, alpha = halpha) +    
    geom_line(aes(y = q95, group = date),
        color=colpal2$q90, lty = 4, lwd = 0.5, alpha = halpha) +
    geom_line(aes(y = q125, group = date),
        color=colpal2$q75, lty = 2, lwd = 0.5, alpha = halpha) +    
    geom_line(aes(y = q875, group = date),
        color=colpal2$q75, lty = 2, lwd = 0.5, alpha = halpha) +
    geom_line(aes(y = q25, group = date),
        color=colpal2$q50, lty = 1, lwd = 0.5, alpha = halpha) +    
    geom_line(aes(y = q75, group = date),
        color=colpal2$q50, lty = 1, lwd = 0.5, alpha = halpha) +
    ## today line
    geom_line(aes(y = q5), data = Qhist %>% filter(date == today()),
        color="black", lty = 4, lwd = 1) +    
    geom_line(aes(y = q95), data = Qhist %>% filter(date == today()),
        color="black", lty = 4, lwd = 1) +
    geom_line(aes(y = q125), data = Qhist %>% filter(date == today()),
        color="black", lty = 2, lwd = 1) +    
    geom_line(aes(y = q875), data = Qhist %>% filter(date == today()),
        color="black", lty = 2, lwd = 1) +
    geom_line(aes(y = q25), data = Qhist %>% filter(date == today()),
        color="black", lty = 1, lwd = 1) +    
    geom_line(aes(y = q75), data = Qhist %>% filter(date == today()),
        color="black", lty = 1, lwd = 1) +
    # geom_linerange(aes(ymin = q5, ymax = q95), 
    #     data = Qhist %>% filter(date == today()),
    #     color = colpal2$q90, lwd = 2) + 
    # geom_linerange(aes(ymin = q125, ymax = q875), 
    #     data = Qhist %>% filter(date == today()),
    #     color = colpal2$q75, lwd = 2) + 
    # geom_linerange(aes(ymin = q25, ymax = q75), 
    #     data = Qhist %>% filter(date == today()),
    #     color = colpal2$q50, lwd = 2) + 
    # geom_hline(yintercept = 0, color='white', lwd = 0.3) +
    xlab("Time (hour)") + ylab("Delay (minutes)") + ylim(-15, 30)
```

* coloured lines represent the
<span class='q90'>5% and 95%</span>,
<span class='q75'>12.5% and 87.5%</span>, and
<span class='q50'>25% and 75%</span> quantiles of bus delays 


## This week vs. last 4 weeks

```{r mmsofarQ, fig.height=5.5, fig.width=9}
mmq <- Qhist %>%
    filter(date >= Monday - weeks(4) & date <= today()) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday', 'Saturday', 'Sunday')),
           week = (format(date, '%W') %>% as.numeric) - format(Monday, '%W') %>% as.numeric) %>%
    select(zero, q5, q125, q25, q75, q875, q95, time, date, dow, week) %>%
    mutate(week = factor(week, levels = -4:-0,
                         labels = c(paste(4:2, 'weeks ago'), "Last week", "This week")))

ggplot(mmq, aes(x = time)) +
    geom_linerange(aes(ymin = q5, ymax = q95), color = colpal2$q90, lwd = 1) + 
    geom_linerange(aes(ymin = q125, ymax = q875), color = colpal2$q75, lwd = 1) + 
    geom_linerange(aes(ymin = q25, ymax = q75), color = colpal2$q50, lwd = 1) + 
    facet_grid(week~dow, drop = FALSE) +
    xlab("Time (hour)") + ylab("Delay (minutes)") + ylim(-15, 30)
```

* the regions contain
<span class='q90'>90%</span>,
<span class='q75'>75%</span>, and
<span class='q50'>50%</span> of buses



<!-- 
View the slides here:

<span class='bigN'>bit.ly/madmarch2018</span>
 -->
<script>
window.onload = function(e) {
    var timer;
    Reveal.addEventListener('slidechanged', function( event ) {
        if (1===2 && Reveal.isLastSlide()) {
            timer = setTimeout(function() {
                window.location = 'http://tomelliott.co.nz/realtimebuses/madmarch.html';
            }, 20000);
        } else {
            clearTimeout(timer);
        }
    });
};
</script>
