---
title: "March Madness"
author: "Tracking Auckland's bus delays<br>through its busiest month"
date: "Based on historical and real-time data, updated approximately every 30 minutes,<br>obtained from Auckland Transport's public GTFS API"
output: 
    revealjs::revealjs_presentation:
        css: march.css
        reveal_options:
            controls: false
            loop: true
            autoSlide: 30000
---

```{r,echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE,error=FALSE)
```

```{r setup,results='hide',message=FALSE}
source('makeplots.R')
load('~/Dropbox/gtfs/delayhistory.rda')
```

## What is "March Madness"?

- Busiest month of the year for public transport
- Buses, trains at capacity
- __Industrial action__: fewer trains during peak, which could increase
bus demand
- Starts February 26, when students surge into the CBD

- __The data__: 
    - real-time delays from buses
    - historical data from April through November 2017 for comparison


## What does a typical day look like?

```{r ngraph,fig.height=3,fig.width=7}
Nhist %>% filter(date == "2018-02-19") %>%
    ggplot(aes(x = time)) + 
        geom_area(aes(y = vlate), fill = "#d35400") +
        # geom_area(aes(y = late), fill = "orange") +
        geom_area(aes(y = ontime), fill = "#1c9c1c") + 
        geom_area(aes(y = early), fill = "#6387ef") +
        # geom_area(aes(y = vearly), fill = "#bb0000") +
        xlim(5, 24) +
        xlab("Time (hour)") + ylab("Number of buses") +
        geom_label(aes(x = 8.25, y = 950, label = 'A')) +
        geom_label(aes(x = 15.75, y = 850, label = 'B')) +
        geom_label(aes(x = 17.75, y = 850, label = 'C'))
```

* Peak periods: __A__ morning, __B__ school, __C__ evening
* The coloured areas represent the proportion<br> of buses that are
    - <span class='early'>more than 5 minutes early</span>
    - <span class='ontime'>on time (within 5 minutes of schedule)</span>
    - <span class='late'>late by more than 5 minutes</span>


```{r latest_ontime}
Nlatest <- Nhist[nrow(Nhist), ]
Qlatest <- Qhist[nrow(Qhist), ]
tcur <- Nlatest$time
tlatest <- paste0(floor(tcur) - ifelse(tcur >= 13, 12, 0), 
                  ":", sprintf("%02d", round((tcur %% 1) * 60)), " ",
                  ifelse(floor(tcur) < 12, "am", "pm"))
total <- Nlatest$vlate
early <- Nlatest$early
ontime <- Nlatest$ontime - early
late <- total - ontime - early
```
## As of `r tlatest` today


* <span class='bigN ontime'>`r ontime`</span> of <span class='bigN total'>`r total`</span> buses were on time
* <span class='bigN late'>`r late`</span> were running more than 5 minutes late
* <span class='bigN early'>`r early`</span> were more than 5 minutes early

```{r todaytrace, message=FALSE, warning=FALSE, error=FALSE, fig.height=3,fig.width=10}
Nday <- Nhist %>% 
    filter(Nhist$date == Sys.Date())
gridExtra::grid.arrange(
    ggplot(Nday, aes(x = time)) + 
        geom_area(aes(y = vlate), fill = "#d35400") +
        geom_area(aes(y = ontime), fill = "#1c9c1c") + 
        geom_area(aes(y = early), fill = "#6387ef") +
        xlim(5, 24) +
        xlab("Time (hour)") + ylab("Number of buses"),
    ggplot(Nday, aes(x = time)) + 
        geom_path(aes(y = (vlate - ontime) / vlate * 100),
            color = "#d35400", lwd = 2) +
        geom_path(aes(y = (ontime - early) / vlate * 100),
            color = "#1c9c1c", lwd = 2) + 
        geom_path(aes(y = early / vlate * 100),
            color = "#6387ef", lwd = 2) +
        ylim(0, 100) + xlim(5, 24) +
        xlab("Time (hour)") + ylab("Percent of buses"),
    ncol = 2)
```

## How does this compare?

* data from April through November 2017
* average percentage of buses <strong class='early'>early</strong>, 
  <strong class='ontime'>ontime</strong>, or <strong class='late'>late</strong>
```{r averages, message=FALSE, warning=FALSE, error=FALSE,fig.height=4,fig.width=12}
Nyear <- Nhist %>%
    filter(date >= '2017-04-01' & date < '2017-12-01' & 
           !format(date, '%a') %in% c('Sat', 'Sun')) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday')))
## Filter out Mon-Fri with max less than 400
dateMax <- with(Nyear, tapply(vlate, date, max))
Nyear <- Nyear %>%
    filter(dateMax[Nyear$date %>% as.character] > 730)
Nlatest$dow <- factor(format(Nlatest$date, "%A"), levels = levels(Nyear$dow))
Nday$dow <- factor(format(Nday$date, "%A"), levels = levels(Nyear$dow))
ggplot(Nyear, aes(x = time)) + 
    # geom_point(aes(y = vlate), color = "#d3540005") + 
    # geom_point(aes(y = ontime), color = "#1c9c1c05") +
    geom_smooth(aes(y = (vlate - ontime) / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#d35400") +
    geom_smooth(aes(y = (ontime - early) / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#1c9c1c") +
    geom_smooth(aes(y = early / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#6387ef") +
    geom_point(aes(y = (vlate - ontime) / vlate * 100), data = Nlatest, color="#d35400", size = 2, stroke=2, pch = 4) +
    geom_point(aes(y = (ontime - early) / vlate * 100), data = Nlatest, color="#1c9c1c", size = 2, stroke=2, pch = 4) +
    geom_point(aes(y = early / vlate * 100), data = Nlatest, color="#6387ef", size = 2, stroke=2, pch = 4) +
    # geom_path(aes(y = vlate), data = Nday, color = "#d35400", lty = 2) +
    # geom_path(aes(y = ontime), data = Nday, color = "#1c9c1c", lty = 2) + 
    # geom_path(aes(y = early), data = Nday, color = "#6387ef", lty = 2) +
    facet_grid(~dow) +
    xlab("Time (hour)") + ylab("Percent of buses") + ylim(0, 100)
```
* crosses show current values (as of `r tlatest`)


## March Madness so far: % on-time

```{r mmsofar, message=FALSE, warning=FALSE, error=FALSE,fig.height=5,fig.width=9}
Nyearavg <- Nyear %>%
    mutate(week = 0)
mm <- Nhist %>%
    filter(date >= '2018-02-26' & date < '2018-04-01' &
           !format(date, '%a') %in% c('Sat', 'Sun')) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday')),
           week = (format(date, '%W') %>% as.numeric) - 8) %>%
    select(zero, vearly, early, ontime, late, vlate, time, dow, week) %>%
    # bind_rows(Nyearavg) %>%
    mutate(week = factor(week, levels = 1:5, 
                        labels = paste('Week', 1:5)))
Nyearavg2 <- Nyearavg %>% 
    select(zero, vearly, early, ontime, late, vlate, time, dow)
ggplot(mm, aes(x = time)) +
    geom_path(aes(y = (vlate - ontime) / vlate * 100), lwd = 1,
        color = "#d35400", data = mm) + # %>% filter(week != "2017 Average")) +
    geom_path(aes(y = (ontime - early) / vlate * 100), lwd = 1,
        color = "#1c9c1c", data = mm) + # %>% filter(week != "2017 Average")) + 
    geom_path(aes(y = early / vlate * 100), lwd = 1,
        color = "#6387ef", data = mm) + # %>% filter(week != "2017 Average")) +
    # geom_smooth(aes(y = (vlate - ontime) / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#d35400") +
    # geom_smooth(aes(y = (ontime - early) / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#1c9c1c") +
    # geom_smooth(aes(y = early / vlate * 100), span = 0.1, method='loess', se=FALSE, color="#6387ef") +
    geom_smooth(aes(y = (vlate - ontime) / vlate * 100), data = Nyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#d3540070", lty = 2, lwd = 0.5) +
    geom_smooth(aes(y = (ontime - early) / vlate * 100), data = Nyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#1c9c1c70", lty = 2, lwd = 0.5) +
    geom_smooth(aes(y = early / vlate * 100), data = Nyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#6387ef70", lty = 2, lwd = 0.5) +
    facet_grid(week~dow, drop = FALSE) +
    xlab("Time (hour)") + ylab("Percent of buses") + ylim(0, 100)
```

* percentage of buses <strong class='early'>early</strong>, 
<strong class='ontime'>ontime</strong>, or <strong class='late'>late</strong>



## But just _how_ late (or early)?

```{r}
library(RProtoBuf)
readProtoFiles(dir = "../assets/protobuf")
pb <- read(transit_realtime.FeedMessage, '~/Dropbox/gtfs/trip_updates.pb')
tobs <- as.POSIXct(pb$header$timestamp, origin = "1970-01-01")
```

* Distribution of vehicle delays at `r format(tobs, '%l:%M %p')`
```{r nowdelays, message=FALSE, warning=FALSE, error=FALSE,fig.height=5,fig.width=10}
delays <- data.frame(
    delay = sapply(pb$entity, function(x) {
        stu <- x$trip_update$stop_time_update[[1]]
        if (stu$has('arrival')) return (stu$arrival$delay)
        return (stu$departure$delay)
    })) %>%
    mutate(g = rep("delay", n()),
           delay = pmax(-20, pmin(30, delay / 60)))
p1 <- ggplot(delays, aes(x = delay)) + 
    geom_dotplot(binwidth = 1/3) + 
    xlim(-20, 30) +
    theme(axis.ticks.y = element_blank(),
          axis.text.y = element_blank()) +
    xlab("Delay (minutes)") + ylab("")
delayq <- data.frame(
        q1 = quantile(delays$delay, 0.05),
        q2 = quantile(delays$delay, 0.125),
        q3 = quantile(delays$delay, 0.25),
        q4 = quantile(delays$delay, 0.75),
        q5 = quantile(delays$delay, 0.875),
        q6 = quantile(delays$delay, 0.95)
    )
p2 <- ggplot(delayq, aes(xmin = -1, xmax = 1)) +
    geom_rect(aes(ymin = q1, ymax = q6), fill = '#bd9218') +
    geom_rect(aes(ymin = q2, ymax = q5), fill = '#914507') +
    geom_rect(aes(ymin = q3, ymax = q4), fill = 'black') +
    coord_flip() +
    ylim(-20, 30) + xlim(-2, 2) +
    theme_void() +
    # theme(axis.ticks.y = element_blank(),
    #       axis.text.y = element_blank(),
    #       axis.ticks.x = element_blank(),
    #       axis.text.x = element_blank()) +
    ylab("") + xlab("")
gridExtra::grid.arrange(p1, p2, heights = c(6, 1))
```
* the brown bar contains 
<span class='q90'>90%</span>,
<span class='q75'>75%</span>, and
<span class='q50'>50%</span> of buses


## March Madness so far: delays

```{r mmsofarQ, message=FALSE, warning=FALSE, error=FALSE,fig.height=5,fig.width=9}
Qyear <- Qhist %>%
    filter(date >= '2017-04-01' & date < '2017-12-01' & 
           !format(date, '%a') %in% c('Sat', 'Sun')) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday')))
## Filter out Mon-Fri with max less than 400
Qyear <- Qyear %>%
    filter(dateMax[Qyear$date %>% as.character] > 730)
Qyearavg <- Qyear %>% mutate(week = 0)
mmq <- Qhist %>%
    filter(date >= '2018-02-26' & date < '2018-04-01' &
           !format(date, '%a') %in% c('Sat', 'Sun')) %>%
    mutate(dow = factor(format(date, '%A'),
                        levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 
                                   'Friday')),
           week = (format(date, '%W') %>% as.numeric) - 8) %>%
    select(zero, q5, q125, q25, q75, q875, q95, time, dow, week) %>%
    mutate(week = factor(week, levels = 1:5, 
                        labels = paste('Week', 1:5)))
Qyearavg2 <- Qyearavg %>% 
    select(zero, q5, q125, q25, q75, q875, q95, time, dow)
ggplot(mmq, aes(x = time)) +
    # geom_ribbon(aes(ymin = q5, ymax = q95), data = Qyearavg2, fill = "#bd921820") + 
    # geom_ribbon(aes(ymin = q125, ymax = q875), data = Qyearavg2, fill = "#91450720") + 
    # geom_ribbon(aes(ymin = q25, ymax = q75), data = Qyearavg2, fill = "#00000020") + 
    geom_smooth(aes(y = q5), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 3, lwd = 0.5) +    
    geom_smooth(aes(y = q95), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 3, lwd = 0.5) +
    geom_smooth(aes(y = q125), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 2, lwd = 0.5) +    
    geom_smooth(aes(y = q875), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 2, lwd = 0.5) +
    geom_smooth(aes(y = q25), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 1, lwd = 0.5) +    
    geom_smooth(aes(y = q75), data = Qyearavg2,
        span = 0.1, method='loess', se=FALSE, color="#2c7aaa", lty = 1, lwd = 0.5) +
    geom_ribbon(aes(ymin = q5, ymax = q95), fill = "#bd9218", alpha=0.8) + 
    geom_ribbon(aes(ymin = q125, ymax = q875), fill = "#914507", alpha=0.8) + 
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#000000", alpha=0.8) + 
    geom_hline(yintercept = 0, color='white', lwd = 0.3) +
    facet_grid(week~dow, drop = FALSE) +
    xlab("Time (hour)") + ylab("Delay (minutes)")
```

* the regions contain
<span class='q90'>90%</span>,
<span class='q75'>75%</span>, and
<span class='q50'>50%</span> of buses
* <span style='color: #2c7aaa'>historical averages</span>

***

View the slides here:

<span class='bigN'>bit.ly/madmarch18</span>

<script>
window.onload = function(e) {
    Reveal.addEventListener('slidechanged', function( event ) {
        if (Reveal.isLastSlide()) {
            setTimeout(function() {
                window.location = 'http://tomelliott.co.nz/realtimebuses/mamarch.html';
            }, 20000);
        }
    });
};
</script>
